Temporality of the Victorian Hospital Patient Transfer Network

### 2025-06-23 reports/diagrams/new-networks/static.tex
Last week's stuff

    \node at (0.0, 0.0) (NORTHWEST){};
    \node at (16.0, 10.0) (SOUTHEAST){};

    \node at (2.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A){$u$};
    \node at (2.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B){$v$};
    \node at (6.0, 5.0) [draw,circle,minimum width=1 cm,inner sep=0](H){$h_{uv}$};

    \node at (10.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A2){$u$};
    \node at (10.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B2){$v$};
    \node at (14.0, 5.0) [draw,circle,minimum width=1 cm,inner sep=0](H2){$h_{uv}$};

    \draw [out=225,in=135,-stealth,solid](A.225)to(B.135);
    \draw [out=0,in=270,-stealth,densely dashed](B.0)to(H.270);

    \draw [out=90,in=0,-stealth,densely dashed](H2.90)to(A2.0);

    \draw [loosely dashed, very thick](8.0, 0.0)--(8.0, 10.0);

>>>>

    \node at (0.0, 0.0) (NORTHWEST){};
    \node at (10.0, 10.0) (SOUTHEAST){};

    \node at (3.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A){$u$};
    \node at (3.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B){$v$};
    \node at (7.0, 5.0) [draw,circle,minimum width=1 cm,inner sep=0](H){$h_{uv}$};

    \draw [out=225,in=135,-stealth,solid](A.225)to(B.135);
    %\draw [out=45,in=315,-stealth,solid](B.45)to(A.315);
    \draw [out=0,in=270,-stealth,densely dashed](B.0)to(H.270);
    \draw [out=90,in=0,-stealth,densely dashed](H.90)to(A.0);

### 2025-06-23 reports/diagrams/new-networks/temporal.tex
Last week's stuff

    \node at (2.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A){$u$};
    \node at (2.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B){$v$};
    \node at (6.0, 5.0) [draw,circle,minimum width=1 cm,inner sep=0](H){$h_{uv}$};

    \node at (10.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A2){$u$};
    \node at (10.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B2){$v$};
    \node at (14.0, 5.0) [draw,circle,minimum width=1 cm,inner sep=0](H2){$h_{uv}$};

    \draw [out=225,in=135,-stealth,solid](A.225)to(B.135);
    \draw [out=0,in=270,-stealth,densely dashed](B.0)to(H.270);

    \draw [out=90,in=0,-stealth,densely dashed](H2.90)to(A2.0);

    \draw [loosely dashed, very thick](8.0, 0.0)--(8.0, 10.0);

>>>>

    \node at (4.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A){$u$};
    \node at (4.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B){$v$};

    \node at (12.0, 8.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](A2){$u$};
    \node at (12.0, 2.0) [draw,solid,minimum width=1 cm,minimum height=1 cm,inner sep=0](B2){$v$};

    \draw [out=225,in=135,-stealth,solid](A.225)to(B.135);
    \draw [out=0,in=180,-stealth,densely dashed](B.0)to(A2.180);

    \draw [loosely dashed,very thick](8.0, 0.0)--(8.0, 10.0);

### 2025-06-23 reports/paper.qmd
Last week's stuff

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from matplotlib import pyplot as plt, dates as mdates, colors as mcolors
import seaborn as sns
import calendar
from itertools import accumulate
from collections import defaultdict
import datetime

@@

@@

In this paper, we investigate these differences using the Victorian hospital system as a baseline for comparison. 
In particular, we investigate the impacts of the inclusion of modelling the period of time that patients spend at home between admissions, and the effect of approximating the granulovement around the hospital system by analysing measures of centrality and clustering.
A weighted, directed network can be used to construct a Markov chain that in turn can be analysed or simulated to understand the dynamics of a spreading infectious disease.
For each edge $e_{ij}$ with a number of observed movements $w_{ij}$ over some time period $\tau$, 
we can construct a reaction that has a hazard function $\lambda_{ij} = w_{ij} / \tau$. This implicitly constructs a process that has inter-event times (i.e. times between distinct patient movements) that are exponentially distributed $\text{Exp}(\lambda_{ij})$.

@@

![Naïve static](diagrams/network_diags/naive_static.pdf)

![Static](diagrams/network_diags/static.pdf)

![Snapshot](diagrams/network_diags/snapshot.pdf)

![Temporal](diagrams/network_diags/temporal.pdf)

Schematic diagrams of each type of network representation for a two-facility network. Solid arrows represent direct transfers, dashed arrows represent indirect transfers, and wavy lines represent temporal consistency.
:::

@@

We plot the (empirical) cumulative distribution function of the hitting times for all pairs of seed facility and target facility in @fig-ecdf

@@

ax.set_xlabel('hitting time')
ax.set_ylabel('ECDF')

@@

One striking feature

```{python}
#| label: compute-load-hosp-presence
histories = {
  dir.stem: np.loadtxt(dir/"hosp_presence.csv", delimiter=',')
  for dir in (root / "simulations/zero_sims_resized/" ).iterdir()
}
```
```{python}
#| label: fig-homeprop-heatmap
#| fig-cap: Distribution of the number of indivdiuals at home (not in the hospital system) over the duration of the simulation for each model type.

_BASE_DATE = np.datetime64('2011-01', 'M')
_MAX_DATE = np.datetime64('2018-12', 'M')
def set_xdatelabels(hist, ax, spacing=12, dt=1.0):
    # here, we want to get succinct date formats
    # we map the times onto the timedeltas of the simulations
    N = hist.shape[1]
    emp_max_date = (np.timedelta64(int(N*dt), 'D') + _BASE_DATE).astype('datetime64[M]')
    ax_max_date = np.min([emp_max_date, _MAX_DATE])

@@

_DTS = defaultdict(lambda: 1.0)
_DTS['temporal'] = 0.5
def single_heatmap(model, hist, bins=(540, 31), norm=None, ax=None, cbar=False):
    if ax is None:
        fig, ax = plt.subplots()

@@

    dt = _DTS[model]
    ts = (np.arange(hist.shape[1]) * dt).reshape((1, -1))
    ts_arr = np.repeat(ts, hist.shape[0], axis=0)

    *_, coll = ax.hist2d(
        ts_arr.flatten(), 
        30 - hist.flatten(),
        bins=bins,
        density=True,
        norm=norm,
    )

@@

_ax_order = {
    'temporal': 0,
    'snapshot': 1,
    'static': 2,
    'naive_static': 3,
}
def final_heatmap(hists, norm=mcolors.PowerNorm(0.5)):

    fig, axs = plt.subplots(nrows=4, figsize=[8.25, 5.25], sharex=True)
    axs_flat = axs.flatten()
    for k, h in hists.items():
        ax = axs_flat[_ax_order[k]]
        coll, _ = single_heatmap(k, h, norm=norm, ax=ax)
        norm = coll.norm
        ax.set_ylabel(k.replace('_', '\n'))
    set_xdatelabels(h, axs_flat[-1])
    fig.subplots_adjust(hspace=0.3)
    fig.colorbar(coll, ax=axs, label='Density', extend='max')
    fig.supylabel("Number at home ($N_0=30$)")

@@

_ = final_heatmap(histories)
```

>>>>

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from matplotlib import pyplot as plt, dates as mdates, colors as mcolors, cm
import seaborn as sns
import calendar
from itertools import accumulate
from collections import defaultdict
import datetime
import copy

@@

Perhaps most pertinently, [@belik_leveraging_2017] explores patient referral networks using a temporal network methodology, and also consider the effect of community stays by augmenting their temporal network with additional edges corresponding to readmissions that occur within a period of $k$ days (representative of the infectious period). Here, we will show that

@@

In this paper, we investigate these differences using the Victorian hospital system as a baseline for comparison. Qualitatively, there are mechanisms implicit in the data that are not previously well-modelled, and may be of importance to constructing robust control measures. Critically, we notice that the distribution of readmission times (the time from a discharge of a given patient from a facility to next admission of that patient at any facility), as plotted in @fig-readmission-time, are not well modelled by an exponential distribution, i.e. poorly approximated by Markov processes.

```{python}
#| label: fig-readmission-time
#| fig-cap: Distribution of empirical readmission times (time from discharge to next admission), truncated at 365 days; and best-fit exponential (via MLE).

readmission_df = pl.read_parquet(root / "data/concordant_networks/readmission_time.parquet")
readmission_df = (readmission_df
  .with_columns(
    Density = pl.col('count') / pl.col('count').sum()
  )
  .rename(
    {
      'readmission_time': 'Readmission Time [days]', 
      'count': 'Count',
    }
  )
)

fig, ax = plt.subplots()

sns.scatterplot(readmission_df, x='Readmission Time [days]', y='Density', ax=ax, label='data')
sns.despine(ax=ax)

ax.set_xscale('asinh')
_xlims = ax.set_xlim(-0.1, None)
ax.set_yscale('log')
_ylims = ax.get_ylim()

# get best fit exponential, we know that gamma = 1/mean
readmission_mean = readmission_df.select(
  pl.col('Readmission Time [days]') * pl.col('Density')
).sum().item()

gamma = 1. / readmission_mean
_xgrid = np.arange(0, 366, 1)
ax.plot(_xgrid, gamma * np.exp(-(_xgrid * gamma)), color='orange', linestyle='dashed', label='best-fit exponential')

ax.set_ylim(*_ylims)
ax.set_xlim(*_xlims)
ax.set_xticks([0, 1, 10, 100])

ax.legend()
```

A weighted, directed network can be used to construct a Markov chain that in turn can be analysed or simulated to understand the dynamics of a spreading infectious disease.
For each edge $e_{ij}$ with a number of observed movements $w_{ij}$ over some time period $\tau$, 
we can construct a reaction that has a hazard function $\lambda_{ij} = w_{ij} / \tau$. This implicitly constructs a process that has inter-event times (i.e. times between distinct patient movements) that are exponentially distributed $\text{Exp}(\lambda_{ij})$.

@@

![Naïve static](diagrams/new-networks/naive_static.pdf)

![Static](diagrams/new-networks/static.pdf)

![Snapshot](diagrams/new-networks/snapshot.pdf)

![Temporal](diagrams/new-networks/temporal.pdf)

Schematic diagrams of each type of network representation for a two-facility network. There exists a direct transfer from $u$ to $v$ and an indirect transfer from $v$ to $u$. 
:::

@@

We plot the (empirical) cumulative distribution function of the hitting times for all pairs of seed facility and target facility in @fig-ecdf, with a zoom of the early-time distribution in @fig-ecdf-close.
We see that the naive static network model produces a hitting time distribution that has a large amount of movement of the marked individuals in a short amount of time, and that the full temporal network model produces much less movement.
This is consistent with what is expected.
We see that the

@@

ax.set_xlabel('Hitting Time (days)')
ax.set_ylabel('ECDF')

```{python}
#| label: fig-ecdf-close
#| fig-cap: Empirical cumulative distributions of the hitting times between pairs of facilities over approxiumately the first year, for different network models.

fig, ax = plt.subplots()
for model in model_order:
    dist = hitting_time_dists[model]
    ax.plot(dist, np.linspace(0, 1, len(dist), endpoint=False), label=model.replace('_', ' '))

ax.set_xlim(-15, 350)
ax.set_ylim(-0.025, 0.15)

ax.set_xlabel('Hitting Time (days)')
ax.set_ylabel('ECDF')

@@

When comparing the aggregate number of individuals present or not present in the hospital system between models, we see some striking differences. The naïve static model has no capability for discharging individuals out of the system, so is significantly different to the other network models.
For the other models, after some burn-in period, they reach a (pseudo-)steady state. 
Notable features:
- the band at N=0 for the snapshot and temporal network models, where some individuals do not discharge - this is usually due to their seed hospital not existing until sometime much later in the simulation. This cannot be modelled with static networks, so instead in the static network, there are bands of lagging that correspond to lower global rates of discharge.
- the differing values of steady state for each model. Temporal model has the most at home, followed by the static model, and then the snapshot model.
- it seems the mechanisms for readmission in the snapshot model lead to periodic bursty readmissions, which results in a lower number of individuals at home.

```{python}
#| label: compute-load-hosp-presence
histories = {
  dir.stem: np.loadtxt(dir/"hosp_presence.csv", delimiter=',')
  for dir in (root / "simulations/zero_sims_resized/" ).iterdir()
}
```
```{python}
#| label: fig-homeprop-heatmap
#| fig-cap: Distribution of the number of individuals at home (not in the hospital system) over the first 3 years of the simulation for each model type.

_BASE_DATE = np.datetime64('2011-01', 'M')
_MAX_DATE = np.datetime64('2018-12', 'M')
def set_xdatelabels(hist, ax, spacing=12, dt=1.0):
    # here, we want to get succinct date formats
    # we map the times onto the timedeltas of the simulations
    N = hist.shape[1]
    emp_max_date = (np.timedelta64(int(N*dt), 'D') + _BASE_DATE).astype('datetime64[M]')
    ax_lim_date = (np.timedelta64(int(np.max(ax.get_xlim())), 'D') + _BASE_DATE).astype('datetime64[M]')
    ax_max_date = np.min([emp_max_date, _MAX_DATE, ax_lim_date])

@@

_DTS = defaultdict(lambda: 1.0)
_DTS['temporal'] = 0.5
def single_heatmap(model, hist, bins=(540, 31), cutoff=None, norm=None, ax=None, cbar=False):
    if ax is None:
        fig, ax = plt.subplots()

@@

    dt = _DTS[model]
    ts = (np.arange(hist.shape[1]) * dt).reshape((1, -1))
    ts_arr = np.repeat(ts, hist.shape[0], axis=0).flatten()

    if cutoff is not None:
      ts_mask = ts_arr < cutoff
      ts_arr = ts_arr[ts_mask]
      hist_arr = hist.flatten()[ts_mask]
    else:
      hist_arr = hist.flatten()

    *_, coll = ax.hist2d(
        ts_arr, 
        30 - hist_arr,
        bins=bins,
        density=True,
        norm=norm,
    )

@@

_ax_order = {
    'temporal': 0,
    'snapshot': 1,
    'static': 2,
    'naive_static': 3,
}
def final_heatmap(hists, norm=mcolors.PowerNorm(0.5), bins=(540, 31), cutoff=None):

    fig, axs = plt.subplots(nrows=4, figsize=[8.25, 5.25], sharex=True)
    axs_flat = axs.flatten()
    for k, h in hists.items():
        ax = axs_flat[_ax_order[k]]
        coll, _ = single_heatmap(k, h, bins=bins, cutoff=cutoff, norm=norm, ax=ax)
        norm = coll.norm
        ax.set_ylabel(k.replace('_', '\n'))
    set_xdatelabels(h, axs_flat[-1])
    fig.subplots_adjust(hspace=0.3)
    coll_norm = copy.deepcopy(coll.norm) 
    coll_norm.autoscale([0.0, bins[0]* coll.norm.vmax])
    scaled_scalar_map = cm.ScalarMappable(norm=coll_norm, cmap=coll.get_cmap())
    fig.colorbar(scaled_scalar_map, ax=axs, label='Density', extend='max')
    fig.supylabel("Number at home ($N_0=30$)")

@@

_ = final_heatmap(histories, norm=mcolors.PowerNorm(0.4), bins=(313,31), cutoff=3*365)
```
